---
title: "SOA Predictive Analytics Exam, Module 3 Sample Project"
output: html_notebook
---

The following chunks provide an opportunity to explore the sample project data.

Run CHUNK 1 to load the data.

```{r}
#CHUNK 1
termLife <- read.csv(file = "TermLife.csv")
head(termLife)
```

Run CHUNK 2 to view some summary statistics. What stands out?

```{r}
#CHUNK 2
summary(termLife)
```

The following items are of interest:
55% have purchased term insurance. So there will be a good opportunity for the data to separate those who purchase from those who don't.
None of the variables have missing value codes. The only immediately suspicious value is an education level of 2.
In some cases zeros are possible (no gifts to charity or an unmarried person having zero for spouse age).
Later on we might want to check to be sure such fields are consistent.

For now, we concentrate on making some graphs.

First, decide which variables are best represented by bar charts. CHUNK 3 makes one for TERM_FLAG. Change the chunk to do the same for other variables. Is there anything interesting?

```{r}
#CHUNK 3

library(ggplot2)
plot.base <- ggplot(data = termLife) + geom_bar()
plot.base + aes(x=TERM_FLAG)
plot.base + aes(x=GENDER)
plot.base + aes(x=AGE)
plot.base + aes(x=MARSTAT)
plot.base + aes(x=EDUCATION)
plot.base + aes(x=SAGE)
plot.base + aes(x=SEDUCATION)
plot.base + aes(x=NUMHH)


```

The variables suitable for bar charts are:
TERM_FLAG - as noted slightly more than half purchased term insurance
GENDER - Fewer than 100 are female. Recall that this refers to who filled out the survey, not who is insured.
MARSTAT - Few have status 2. It might make sense to combine with status 0.
EDUCATION - An arbitrary decision is made here. Both AGE and EDUCATION are discrete, but there are a relatively small number of EDUCATION values, so no need to group them into a histogram.
SEDUCATION - As noted, there should be zeroes for those with no spouse.
NUMHH - Nothing to comment on

CHUNK 4 makes histograms for the continuous variables, starting with FACE. Again, see if there is anything interesting.

```{r}
# CHUNK 4
plot.base <- ggplot(data = termLife) + geom_histogram()
plot.base + aes(x=FACE)
plot.base + aes(x=AGE)
plot.base + aes(x=SAGE)
plot.base + aes(x=INCOME)
plot.base + aes(x=INCOME) + scale_x_log10()
plot.base + aes(x=CHARITY)

plot.base <- ggplot(data = subset(termLife, FACE>0)) + geom_histogram()
plot.base + aes(x=FACE)
plot.base <- ggplot(data = subset(termLife, SAGE>0)) + geom_histogram()
plot.base + aes(x=SAGE)
plot.base <- ggplot(data = subset(termLife, CHARITY>0)) + geom_histogram()
plot.base + aes(x=CHARITY)

```

FACE - With 45% being zero, it might be better to plot only those values where FACE is positive. CHUNK 5 provides this vies.
AGE - Appears to be nicely distributed across reasonable values
SAGE - Ignoring the zeroes, this also seems reasonable
INCOME - Highly skewed, so consider a log scale (code not provided)
CHARITY - There are a lot of zeroes. CHUNK 6 removes them before plotting

```{r}
# CHUNK 5

termLife.pos <- termLife[termLife$FACE>0, ]
plot.base <- ggplot(data = termLife.pos) + geom_histogram()
plot.base + aes(x=FACE)
plot.base + aes(x=FACE) + scale_x_log10()
```

FACE - It is skewed to the right. Consider a log scale for a more useful view (code not provided)

```{r}
#CHUNK 6

termLife.char <- termLife[termLife$CHARITY>0, ]
plot.base <- ggplot(data = termLife.char) + geom_histogram()
plot.base + aes(x=CHARITY)
plot.base + aes(x=CHARITY) + scale_x_log10()
```

CHARITY - It is skewed to the right . Consider a log scale for a more useful view (code not provided)

The last graphs we will make are scatterplots. CHUNK 7 makes one for FACE versus INCOME. Make any others that might be interesting and comment.

```{r}
# CHUNK 7
plot.base.pos <- ggplot(data=termLife.pos,
            mapping = aes(x = INCOME,
                          y = FACE))
plot.base.pos + geom_point() +
  geom_smooth(method = "loess") +
  scale_x_log10() + scale_y_log10()

```

The plot is interesting in that there appears to be a strong positive linear relationship in the middle incomes (where most of the data are) but it flattens out at each end. A plot of FACE against EDUCATION shows little, if any, relationship.

