---
title: 'Predictive Analytics Exam Module 5 Section 3'
output:
  html_notebook: default
  
---

Run CHUNK 1 to load the mortality data and ggplot2.

```{r echo = FALSE}
#CHUNK 1
data.mortality <- read.csv("soa_mortality_data.csv")
library(ggplot2)
```

Run CHUNK 2 to check for outliers in the duration variable.

```{r echo = FALSE}
#CHUNK 2
library(gridExtra)
p1 <- ggplot(data = data.mortality, aes(duration, ..density..)) +
       geom_histogram(binwidth = 1, fill = "blue", alpha = 0.5)
p2 <- ggplot(data = data.mortality, aes(x = "duration", y = duration)) +
        geom_boxplot(fill = "blue", alpha = 0.5) 
grid.arrange(p1,p2,ncol=2)
```

In CHUNK 3 modify the code from CHUNK 2 to check for outliers in the duration variable after a log transformation.

```{r echo = FALSE}
#CHUNK 3
p1 <- ggplot(data = data.mortality, aes(log(duration), ..density..)) +
       geom_histogram(binwidth = 0.1, fill = "blue", alpha = 0.5)
p2 <- ggplot(data = data.mortality, aes(x = "log(duration)", y = log(duration))) +
        geom_boxplot(fill = "blue", alpha = 0.5)
grid.arrange(p1,p2,ncol=2)
```

CHUNK 4 has the modified code.

```{r echo = FALSE}
#CHUNK 4
data.mortality$durationLog <- log(data.mortality$duration)
p1 <- ggplot(data = data.mortality, aes(durationLog, ..density..)) +
       geom_histogram(binwidth = 0.1, fill = "blue", alpha = 0.5) 
p2 <- ggplot(data = data.mortality, aes(x = "duration", y = durationLog)) +
        geom_boxplot(fill = "blue", alpha = 0.5) 
grid.arrange(p1,p2,ncol=2)
```

Run CHUNK 5 to see the effect of an outlier on a regression line.

```{r echo = FALSE}
#CHUNK 5
#The x-values are uniform(0,10) random numbers
#The y-values are the x-values plus a standard normal variable.
sample.x <- runif(10, min = 0, max = 10)
sample.y <- sample.x + rnorm(10, 0, 1)
#The "real" regression line should be rougly the line y=x.
line.real <- lm(sample.y~sample.x)
#Now change the tenth y-value to 50, an outlier
sample.y[10] = 50
#Redo the regression.
line.bias <- lm(sample.y~sample.x)
#Plot the two regression lines.
plot(sample.y~sample.x)
abline(line.real, col = "blue")
abline(line.bias, col = "red")
```

Run CHUNK 6 to revisit the duration variable on a log scale.

```{r echo = FALSE}
#CHUNK 6

p1 <- ggplot(data = data.mortality, aes(x = "duration", y = duration)) +
        geom_boxplot(fill = "orange", alpha = 0.5) +
        scale_y_log10()
p2 <- ggplot(data = data.mortality, aes(x = prodcat, y = duration)) +
        geom_boxplot(fill = "orange", alpha = 0.5) +
        scale_y_log10()
grid.arrange(p1,p2,ncol=2)
```

CHUNK 7 makes a graph that illustrates the difference between squared and absolute error. It is provided here in case you are interested in how the graph was made.

```{r echo = FALSE}
#CHUNK 7

x <- seq(-1.5,1.5,0.1)
errors <- data.frame(error = x,
                     squared = x^2,
                     abs = abs(x))

ggplot(data = errors, aes(x = error)) +
  geom_line(aes(y = squared), color = "red") +
  geom_line(aes(y = abs), color = "blue") +
  geom_line(aes(y = 1)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = 1, fill = "yellow", alpha = 0.2) + 
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1, ymax = Inf, fill = "orange", alpha = 0.2)

```

Run CHUNK 8 to examine duration versus actual face for term insurance products.

```{r}
#CHUNK 8
p1 <- ggplot(data = subset(data.mortality, prodcat == "TRM"), aes(x = duration, y = actual_face)) +
        geom_point(alpha = 0.5) +
        scale_x_continuous(limits = c(0, 80)) + 
        scale_y_log10(limits = c(1, 1.0e+07)) +
        annotate("rect", xmin = 39, xmax = 43, ymin = 600, ymax = 10000, fill = "red", alpha = 0.2)
p1
```

Run CHUNK 9 to see an example where one variable is categorical and the other is numeric.

```{r}
#CHUNK 9
p1 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = smoker, y = issage, fill = smoker)) +
        geom_boxplot() 
p2 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = "smoker", y = issage)) +
        geom_boxplot() 
grid.arrange(p1,p2,ncol=2)
```

Run CHUNK 10 to make a frequency table of distribution channel and smoking status.

```{r}
#CHUNK 10
tbl <- table(data.mortality$distchan, data.mortality$smoker)
tbl
```
