---
title: "Predictive Analtyics Exam Module 6 Section 1"
output:
  html_document:
    df_print: paged
---

Run CHUNK 1 to load some pacakges.

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
#CHUNK 1
library(ggplot2)
library(gridExtra)

```

Run CHUNK 2 to load and plot the Galton data.

```{r Answer - galton plot}
#CHUNK 2
galton <- read.csv("galton.csv")
set.seed(1000)

head(galton)
ggplot(galton, aes(x = jitter(parent, 5), y = jitter(child, 5))) +
  geom_point() +
  labs(x = "Parents' Height", y = "Child's Height")
```

Run CHUNK 3 to plot the regression line.

```{r}
#CHUNK 3
set.seed(1000)
ggplot(galton, aes(x = jitter(parent, 5), y = jitter(child, 5))) +
  geom_point() +
  labs(x = "Parents' Height", y = "Child's Height") +
  stat_smooth(method="lm", col = "red")

```

Run CHUNK 4 to formally fit the line.

```{r model_summary}
#CHUNK 4
lm_galton <- lm(child ~ parent, data = galton)
summary(lm_galton)
```
  
Use CHUNK 5 to work on Exercise 6.1.1

```{r Answer 6.7.1}
#CHUNK 5

```

CHUNK 6 has a solution.

```{r}
#CHUNK 6
b0 <- coef(lm_galton)[1]
b1 <- coef(lm_galton)[2]
x <- 68

# Manual calculation of child height
yhat <- b0 + b1*x
print(yhat)

# Using predict function
yhat2 <- predict(lm_galton, newdata = data.frame(parent = 68))
print(yhat2)
```

Run CHUNK 7 to load and set up the auto claim dataset.

```{r}
#CHUNK 7

AutoClaim <- read.csv("AutoClaim.csv")

# We will only use GENDER, AGE, BLUEBOOK, and CLM_AMT for demonstration
AutoClaim_sub <- subset(AutoClaim, select = c('GENDER','AGE','BLUEBOOK','CLM_AMT'))
head(AutoClaim_sub)
# Create age bands and then remove age
AutoClaim_sub$AGE_BAND <- cut(x = AutoClaim_sub$AGE, breaks = c(0,25,35,45,55,65,85))
AutoClaim_sub$AGE <- NULL
# Select only cases where CLM_AMT is positive
AutoClaim_sub <- AutoClaim_sub[AutoClaim$CLM_AMT > 0,]

head(AutoClaim_sub)
```

Run CHUNK 8 to set up and view part of the model matrix.

```{r}
MM <- model.matrix(CLM_AMT ~ GENDER + AGE_BAND + BLUEBOOK, data = AutoClaim_sub)
MM[1:6,]
```

Run CHUNK 9 to obtain the parameter estimates by matrix algebra.

```{r}
#CHUNK 9
X <- MM
Y <- AutoClaim_sub$CLM_AMT

#R does not have a matrix inverse function, but solve() performs this task.
Beta <- solve(t(X)%*%X) %*% (t(X)%*%Y)
Beta
```

Run CHUNK 10 to obtain the least squares estimates.

```{r lm_model}
linear.model <- lm(CLM_AMT ~ GENDER + AGE_BAND + BLUEBOOK, data = AutoClaim_sub)
linear.model
```

Run CHUNK 11 to see a visualiztion of the difference how conditional normal distributions does not imply an aggregate normal distribution.

```{r warning=FALSE, message=FALSE}
#CHUNK 10
set.seed(42)

f <- rnorm(n = 500, mean = 1500, sd = 400)
m <- rnorm(n = 500, mean = 3000, sd = 400)
gend <- data.frame(CLM_AMT = f, GENDER = 'F')
gend <- rbind(gend, data.frame(CLM_AMT = m, GENDER = 'M'))

p1 <- ggplot(gend, aes(x=CLM_AMT)) + 
        geom_freqpoly(position="identity", col = 'black') + 
        ggtitle(label = "Overall distribution of CLM_AMT")

p2 <- ggplot(gend, aes(x=CLM_AMT, fill=GENDER)) + 
        geom_freqpoly(position="identity", col = 'black') + 
        ggtitle(label = "CLM_AMT distribution by GENDER")

grid.arrange(p1, p2, ncol = 2)
```

Use CHUNK 12 to test the assumptions of ordinary least squares for Exercise 6.1.2.

```{r}
#CHUNK 12

#run the regression
model <- lm(CLM_AMT ~ GENDER + AGE_BAND + BLUEBOOK, data = AutoClaim_sub)

#Check the mean of the residuals
mean(model$residuals) 

#Check the residuals for constant variance. Not all four plots relate to this assumption.
par(mfrow=c(2,2))
plot(model)

#Check that the residuals and the predictor variables are uncorrelated
cor.test(AutoClaim_sub$BLUEBOOK, model$residuals) # Run this for each your predictor variables

#Check that the residuals have a normal distribution
#One check is a Q-Q plot, which appears in the upper right corner of the plots made when checking for constant variance. Another option is to make a histogram of the residuals.
resid <- data.frame(model$residuals)
ggplot(resid, aes(x=model.residuals)) + 
        geom_histogram(bins=100, position="identity", col = 'black')

```

